Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-07-26 paper/paper.Rmd
Declare C as a function of p, but it is fixed for given y_i since p=y_i

Motivated by the estimator in \@ref(eq:denriem), we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold learning embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$, and $C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)$ is a correction term for the volume between the spaces so that the integral is 1. We will now state below how to calculate each term in the estimator. And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.

@@

As pointed out by @Perrault-Joncas2013-pq, the naive distance $\|\pmb{p}-\pmb{q}\|$ underestimates $d_g$, and the graph shortest distance $d_G$ overestimates $d_g$ by the triangle inequality. By using only the first-order correction in $\hat{d}_g$, our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in $\pmb{H}(\pmb{y}_i)$, we further divide the density estimator in \@ref(eq:denestimator) by a constant term $C$ so that the integral of the kernel is equal to 1. The constant term in $1$-d is given by the *area under the curve (AUC)*, and can be named as the *volume between the surface (VBS)* in higher dimension. We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and get
\begin{equation}
\label{eq:vbs}
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
\end{equation}
where the distortion is corrected for the tangent plane of the embedding points.

>>>>

Motivated by the estimator in \@ref(eq:denriem), we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold learning embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$, and $C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)$ is a correction term for the volume between the surface when estimating over the sample points so that the integral of the density is 1. We will now state below how to calculate each term in the estimator. And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.

@@

As pointed out by @Perrault-Joncas2013-pq, the naive distance $\|\pmb{p}-\pmb{q}\|$ underestimates $d_g$, and the graph shortest distance $d_G$ overestimates $d_g$ by the triangle inequality. By using only the first-order correction in $\hat{d}_g$, our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in $\pmb{H}(\pmb{y}_i)$, we further divide the density estimator in \@ref(eq:denestimator) by a constant term $C$ so that the integral of the kernel is equal to 1. The constant term in $1$-d is given by the *area under the curve (AUC)*, and can be named as the *volume between the surface (VBS)* in higher dimension. We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and the estimated point $\pmb{p}$ so that for $\pmb{Z} = \{\pmb{y}_1, \dots, \pmb{y}_N, \pmb{p}\}$, we have
\begin{equation}
\label{eq:vbs}
C(\pmb{p}) = \sum_{i=1}^{N+1} \sqrt{\det(\pmb{H}(\pmb{z}_i))} \Delta z^1(\pmb{z}_i)\dots \Delta z^d(\pmb{z}_i),
\end{equation}
where $\pmb{z}_i=(z^1(\pmb{z}_i),\dots, z^d(\pmb{z}_i))$. The distortions are corrected for the tangent plane of the embedding points. Since we aim to estimate the density of the manifold for the embedding of sample points, $\pmb{p}$ is equal to $\pmb{y}_i$, for $i=1,\dots,N$. Therefore, we have
\begin{equation}
\label{eq:vbsyi}
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i).
\end{equation}

### 2022-07-26 paper/paper.tex
Declare C as a function of p, but it is fixed for given y_i since p=y_i

Motivated by the estimator in \eqref{eq:denriem}, we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold learning embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)\) is a correction term for the volume between the spaces so that the integral is 1. We will now state below how to calculate each term in the estimator. And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.

@@

As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|\pmb{p}-\pmb{q}\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality. By using only the first-order correction in \(\hat{d}_g\), our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in \(\pmb{H}(\pmb{y}_i)\), we further divide the density estimator in \eqref{eq:denestimator} by a constant term \(C\) so that the integral of the kernel is equal to 1. The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface (VBS)} in higher dimension. We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and get
\begin{equation}
\label{eq:vbs}
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
\end{equation}
where the distortion is corrected for the tangent plane of the embedding points.

>>>>

Motivated by the estimator in \eqref{eq:denriem}, we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of \(f\), same as \eqref{eq:denestimator0}, based on the manifold learning embedding \(\pmb{y}_i, i = 1, 2, \dots, N\),
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where \(r\) is a global bandwidth scalar, \(\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N\), is the pointwise variable Riemannian matrix and it is positive definite with dimension \(d\times d\), \(\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}\) is the estimated volume density function when using the coordinates from any manifold learning mapping, \(\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)\) is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric \(\pmb{H}(\pmb{y}_i)\), and \(C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)\) is a correction term for the volume between the surface when estimating over the sample points so that the integral of the density is 1. We will now state below how to calculate each term in the estimator. And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.

@@

As pointed out by \textcite{Perrault-Joncas2013-pq}, the naive distance \(\|\pmb{p}-\pmb{q}\|\) underestimates \(d_g\), and the graph shortest distance \(d_G\) overestimates \(d_g\) by the triangle inequality. By using only the first-order correction in \(\hat{d}_g\), our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in \(\pmb{H}(\pmb{y}_i)\), we further divide the density estimator in \eqref{eq:denestimator} by a constant term \(C\) so that the integral of the kernel is equal to 1. The constant term in \(1\)-d is given by the \emph{area under the curve (AUC)}, and can be named as the \emph{volume between the surface (VBS)} in higher dimension. We use the volume estimator in Section 6.4 of \textcite{Perrault-Joncas2013-pq} to sum over all embedding points \(\pmb{y}_i\) and the estimated point \(\pmb{p}\) so that for \(\pmb{Z} = \{\pmb{y}_1, \dots, \pmb{y}_N, \pmb{p}\}\), we have
\begin{equation}
\label{eq:vbs}
C(\pmb{p}) = \sum_{i=1}^{N+1} \sqrt{\det(\pmb{H}(\pmb{z}_i))} \Delta z^1(\pmb{z}_i)\dots \Delta z^d(\pmb{z}_i),
\end{equation}
where \(\pmb{z}_i=(z^1(\pmb{z}_i),\dots, z^d(\pmb{z}_i))\). The distortions are corrected for the tangent plane of the embedding points. Since we aim to estimate the density of the manifold for the embedding of sample points, \(\pmb{p}\) is equal to \(\pmb{y}_i\), for \(i=1,\dots,N\). Therefore, we have
\begin{equation}
\label{eq:vbsyi}
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i).
\end{equation}
